---
title: "Where do Scientists Put their Money in Politics?"
subtitle: "Analyzing Data from FiveThirtyEight's Blog"
author: "Benjamin Ackerman"
date: "February 18, 2018"
output: html_document
---

<style>
  .col2 {
    columns: 2 200px;         /* number of columns and width in pixels*/
    -webkit-columns: 2 200px; /* chrome, safari */
    -moz-columns: 2 200px;    /* firefox */
  }
  .col3 {
    columns: 3 100px;
    -webkit-columns: 3 100px;
    -moz-columns: 3 100px;
  }
</style>


Since FiveThirtyEight recently [shared access to a lot of their data](https://data.fivethirtyeight.com), I thought it would be fun to download and analyze data from one of their blog posts in R.  I remember reading this article on [scientists' donations to political campaigns](https://fivethirtyeight.com/features/when-scientists-donate-to-politicians-its-usually-to-democrats/), and since the original blog post does not feature any visualizations, I thought I'd come up with some of my own.

Here are the packages I will be using:
```{r setup, echo=TRUE,results='hide', message=FALSE}
library(RCurl)
library(reshape)
library(dplyr)
library(ggplot2)
library(stringr)
library(rebus)
library(maps)
library(leaflet)
library(tigris)
library(purrr)
library(knitr)
library(kableExtra)
```

### Download and clean data from 538's website
```{r download and clean data,cache=TRUE}
link = "https://media.githubusercontent.com/media/fivethirtyeight/data/master/science-giving/science_federal_giving.csv"
data = read.csv(text = getURL(link),stringsAsFactors = FALSE)

### Clean up missingness in the data
data_cleaned = data %>% 
  mutate_all(funs(ifelse(. %in% c("","NULL"),NA,.)))
```

### Summarise donations to 2016 presidential candidates by science classification group
```{r presidential election donations, fig.align='center',fig.width = 10,fig.height = 8}
president_funding = data_cleaned %>% 
  mutate(transaction_amt = as.numeric(transaction_amt)) %>% 
  filter(cand_pty_affiliation %in% c("DEM","REP","LIB","IND"),
         cycle == "2016",
         cand_office == "P",
         cand_status == "C") %>% 
  group_by(state,cand_pty_affiliation,cand_name,classification) %>% 
  summarise(money = sum(transaction_amt,na.rm=TRUE))

### See the top 10 candidates who received funding from scientists by classification
top_10_funded = president_funding %>% 
  group_by(cand_name) %>% 
  summarise(total_donations = sum(money)) %>% 
  ungroup() %>% 
  top_n(10,wt = total_donations)

president_funding %>% 
  group_by(cand_name,classification) %>% 
  summarise(mon = sum(money)/1000000) %>% 
  right_join(top_10_funded,by="cand_name")%>%
  ggplot()+
    geom_bar(aes(x = reorder(cand_name,mon),y=mon,fill=classification),stat='identity') + 
    coord_flip() +
    scale_y_continuous(expand=c(0,0)) + 
    theme(legend.position = c(0.8, 0.15),legend.background = element_rect(fill="white")) +
    labs(title="Campaign Donations by Scientists to 2016 Presidential Candidates", x = "Presidential Candidate",y = "Total Donations Received ($, millions)")
```

### Visualize Differences in Scientists' Donations to Congressional Races in 2016
The FiveThirtyEight article noted that scientists were more willing to show support to candidates with scientific backgrounds.  Just for validation, we can look at a list of the ten Congressional Candidates in 2016 who received the most donations from scientists:
```{r congress races, eval = FALSE, echo= FALSE, fig.align = 'center'}
congress_map = us_congressional() %>% 
  mutate(cand_office_district = str_replace(cd115fp, START %R% "0",""),
         state_name = tolower(state_name)) %>% 
  dplyr::rename(cand_office_st = state_abbr) %>% 
  select(cand_office_st,cand_office_district,state_name,geometry)

congress_funding = data_cleaned %>% 
  mutate(transaction_amt = as.numeric(transaction_amt)) %>% 
  filter(cand_pty_affiliation %in% c("DEM","REP"),
         cycle == "2016",
         cand_office == "H",
         cand_status == "C")%>% 
  group_by(cand_office_st,cand_pty_affiliation,cand_office_district) %>% 
  summarise(money = sum(transaction_amt,na.rm=TRUE)) %>% 
  cast(cand_office_st + cand_office_district ~ cand_pty_affiliation, value="money") %>% 
  filter(!is.na(cand_office_district)) %>% 
  mutate_all(funs(ifelse(is.na(.),0,.))) %>% 
  mutate(party_diff = DEM - REP,
         cand_office_district = ifelse(cand_office_district == 0, 1, cand_office_district)) %>% 
  right_join(congress_map,by=c("cand_office_st","cand_office_district"))


congress_donations <- function(region){
  if(region == "usa"){
    return(
      congress_funding %>% 
           filter(!cand_office_st %in% c("HI","AK")) %>% 
           ggplot() + 
              geom_sf(aes(fill=party_diff),colour="grey",size=.5) +
              scale_fill_gradient2("Donation Diffs ($)",low = "darkred", mid = "white", high = "darkblue")+
              theme_void()+
              theme(panel.grid.major = element_line(colour = "white"))+
              labs(title = "Partisan Differences in Scientist Donations to 2016 Congressional Races")+coord_equal())
  }
  else{
  return(
    congress_funding %>% 
      filter(state_name %in% region) %>% 
      ggplot() + 
        geom_sf(aes(fill=party_diff),colour="grey") +
        scale_fill_gradient2("Donation Diffs ($)",low = "darkred", mid = "white", high = "darkblue") +
        theme_void() + coord_equal()+
        theme(panel.grid.major = element_line(colour = "white")) +
        labs(title = "Partisan Differences in Scientist Donations to 2016 Congressional Races", subtitle = region))}
}


ggplotly(congress_donations('virginia'),tooltip = c("fill"))

ggplotly(congress_donations("usa"))

congress_donations("virginia")

congress_donations("north carolina")

g = purrr::map(c("california","virginia","north carolina","texas"),
           function(x) {
             ggplot() + 
               geom_sf(data = filter(congress_funding, state_name == x), aes(fill = party_diff)) +
               theme_void()+scale_fill_gradient2("Donation Diffs ($)",low = "darkred", mid = "white", high = "darkblue") + 
               theme(panel.grid.major = element_line(colour = "white")) + 
               labs(subtitle = x)
           })

g2 <- cowplot::plot_grid(plotlist = g)
g2
```


```{r congress races most donations, eval=TRUE, echo=TRUE, results='asis'}
data_cleaned %>% 
  mutate(transaction_amt = as.numeric(transaction_amt),
         district = paste(cand_office_st,cand_office_district,sep="-")) %>% 
  filter(cand_pty_affiliation %in% c("DEM","REP"),
         cycle == "2016",
         cand_office == "H",
         cand_status == "C")%>% 
  group_by(district,cand_pty_affiliation,cand_name) %>% 
  summarise(donations = sum(transaction_amt, na.rm=TRUE)) %>% #,
            # pct_out_of_state = sum(cand_office_st != state, na.rm = TRUE)/n() * 100)
  ungroup() %>% 
  arrange(desc(donations)) %>% 
  head(10) %>% 
  kable("html") %>% 
  kable_styling()
```

Of the list above, the two with the greatest amount of donations from scientists were Bill Foster, a former physicist, and Shaughnessy Naughton, a former drug researcher, both of whom were referenced in the original article.  Also on the list were Dr. Dena Grayson (MD, PhD), Raja Krishnamoorthi (Bachelor's degree in Mechanical Engineering), John Plumb (MS in Physics), and Jerry McNerney (PhD in Mathematics). While not all of these candidates won their respective races, several continue to dedicate time and resources to encouraging other scientists to [run for office](http://www.314action.org/mission-1/).

Next, I 

```{r congress races updated, eval = TRUE, cache=TRUE,message=FALSE,results='hide'}
all_funding_data  = readRDS("all_funding_data.RDS")

congress_funding = data_cleaned %>% 
  mutate(transaction_amt = as.numeric(transaction_amt)) %>% 
  filter(cand_pty_affiliation %in% c("DEM","REP"),
         cycle == "2016",
         cand_office == "H",
         cand_status == "C")%>% 
  group_by(cand_office_st,cand_pty_affiliation,cand_office_district) %>% 
  summarise(money = sum(transaction_amt,na.rm=TRUE)) %>% 
  cast(cand_office_st + cand_office_district ~ cand_pty_affiliation, value="money") %>% 
  filter(!is.na(cand_office_district)) %>% 
  mutate_all(funs(ifelse(is.na(.),0,.))) %>% 
  mutate(party_diff = DEM - REP) %>% 
  left_join(state.fips %>% 
              select(cand_office_st = abb, fips) %>% 
              filter(!duplicated(fips)),by="cand_office_st") %>% 
  mutate(fips = ifelse(nchar(fips)== 1, paste0("0",fips),fips),
         cand_office_district = ifelse(nchar(cand_office_district)== 1, paste0("0",cand_office_district),cand_office_district),
         ID = paste(fips,cand_office_district,sep="_"),
         cand_office_district = str_replace(cand_office_district,"00","01"),
         ID_name = paste0(cand_office_st,cand_office_district)) %>% 
  left_join(all_funding_data,by=c("cand_office_st","cand_office_district")) %>% 
  rowwise() %>% 
  mutate(dem_pct = ifelse(!D == 0, DEM/D * 100, 0),
         rep_pct = ifelse(!R == 0, REP/R * 100, 0),
         partisan_diff = sum(dem_pct,-rep_pct,na.rm=TRUE))
```

While it may be interesting to look at the candidates who received the most donations from scientists, I was also really curious to determine where donations from scientists were most ``impactful." By combining data from FiveThirtyEight with information on the total amount of money raised by each Congressional Candidate in the 2016 cycle, I wanted to see what percentage of the donations by political party and congressional district came from scientists.

<div class="col2">
```{r top percentages, echo=FALSE}
congress_funding %>% 
  arrange(desc(dem_pct)) %>% 
  mutate(district = paste(cand_office_st,cand_office_district,sep="-")) %>% 
  select(district,percent = dem_pct,winner) %>% 
  head(5) %>% 
  kable("html",caption = "Percent of Total Donations to Democratic Candidates from Scientists")%>% 
  kable_styling()

congress_funding %>% 
  arrange(desc(rep_pct)) %>% 
  mutate(district = paste(cand_office_st,cand_office_district,sep="-")) %>% 
  select(district,percent = rep_pct,winner) %>% 
  head(5) %>% 
  kable("html",caption = "Percent of Total Donations to Republican Candidates from Scientists") %>% 
  kable_styling()
```
</div>

```{r plot setup, echo=FALSE,message=FALSE, results='hide'}
districts = congressional_districts(cb=TRUE,year = 2016)
districts$ID = paste(districts$STATEFP,districts$CD115FP,sep="_")

data = geo_join(districts, congress_funding, 'ID', 'ID', how = 'left')

dom = c(min(congress_funding$party_diff),-min(congress_funding$party_diff))

test = colorRampPalette(c("red","white","blue"))
pal2 <- colorBin(test(20), domain = dom, bins = 20,na.color=NA)

data1 = data[which(data$party_diff < max(dom)),]
data2 = data[which(data$party_diff > max(dom)),]

popup1 <- paste(sep = "<br/>", 
               paste0("<b>",data1$cand_office_st,"-",data1$cand_office_district,": ", data1$winner,"</b>"),
               #paste0("<img src='http://www.house.state.tx.us/photos/members/", txlege$jpg, "' />"), 
               paste0("<b>Donations from Scientists to Dems: </b> $", data1$DEM, " (",round(data1$dem_pct,2),"% of total)"), 
               paste0("<b>Donations from Scientists to Reps: </b> $", data1$REP, " (",round(data1$rep_pct,2),"% of total)"))

popup2 <- paste(sep = "<br/>", 
               paste0("<b>",data2$cand_office_st,"-",data2$cand_office_district,": ", data2$winner,"</b>"),
               #paste0("<img src='http://www.house.state.tx.us/photos/members/", txlege$jpg, "' />"), 
               paste0("<b>Donations from Scientists to Dems: </b> $", data2$DEM, " (",round(data2$dem_pct,2),"% of total)"), 
               paste0("<b>Donations from Scientists to Reps: </b> $", data2$REP, " (",round(data2$rep_pct,2),"% of total)"))
```

```{r plot congress donations, echo=FALSE, eval = TRUE, fig.align = "center",warning=FALSE,message=FALSE}
leaflet() %>%
  setView(lng = -98.5, lat =39.5,zoom=4) %>% 
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = data1, 
              fillColor = ~colorBin(test(20), domain = dom, bins = 20,na.color=NA)(party_diff),
              opacity=1,
              weight = 1,
              color = "white",
              dashArray = "3",
              popup = popup1,
              fillOpacity = .9) %>% 
  addPolygons(data = data2, 
              fillColor = "darkblue",
              opacity=1,
              weight = 1,
              color = "white",
              dashArray = "3",
              popup = popup2,
              fillOpacity = .9) %>% 
  addLegend(position = "bottomleft",
            colors = c("blue", "red"), 
            labels = c("Democratic Candidates", "Republican Candidates"), 
            opacity = 1, 
            title = "More Donations by Scientists to:")
```